# Data Preparation {#data_prep}

```{r, include=FALSE, warning=F, message=F}
# data mgmt
library(tidyverse)
library(lubridate)
# visualization
library(scales)
library(viridis)
library(kableExtra)
library(latex2exp)
# spatial
library(sf)
library(terra)
library(stars)
library(starsExtra) # functions for working with stars objects
library(lwgeom) 
library(mapview) #Interactive maps
library(leafpop) #map html popup
library(pracma) # Practical Numerical Math Functions

# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  , results='hide'
  , fig.width = 10
  , fig.height = 7
)
```

```{r}
# turn off the s2 processing 
## https://stackoverflow.com/questions/68478179/how-to-resolve-spherical-geometry-failures-when-joining-spatial-data
sf::sf_use_s2(FALSE)

# set crs
my_crs <- 5070 # EPSG:5070 = NAD83/Conus Albers (units = meters); EPSG:4326 = WGS 84 (units = dd); EPSG:4269 = NAD83 (units = dd)
```

## Obtain Elevation Data

The raster [elevation data](https://drive.google.com/drive/folders/1zpf4jUtbMu5shYyqqTMieX0oydI0SP6E?usp=sharing) sent by Mark Melham (Mark.Melham@colostate.edu) was in a geodatabase (.gdb). GDAL uses the OpenFileGDB driver, which can only read vector data sources ([here](https://gis.stackexchange.com/questions/267186/importing-raster-with-r-from-file-geodatabase)). Have to open .gdb in Arc and export as GeoTIFF (.tif).

```{r, eval=FALSE}
# read in elevation data exported from Arc as .tif
elev_temp <- terra::rast("../data/ground_elev_reproj.tif")
# check values
terra::summary(elev_temp)
# appears that elevation is in ft...convert to m
elev_temp <- elev_temp/3.2808
# map
elev_temp |> 
  terra::aggregate(fact=10, fun="mean") |> 
  as.data.frame(xy=T) |> 
  rename(elevation=3) |> 
  ggplot() +
    geom_tile(mapping = aes(x=x,y=y,fill=elevation)) +
    scale_fill_viridis_c(option = "cividis", label = scales::comma) +
    scale_x_continuous(expand = c(0.01, 0.01)) +
    scale_y_continuous(expand = c(0.01, 0.01)) +
    labs(
      fill = "Elevation (m)"
    ) +
    theme_light() +
    theme(
      legend.position = "bottom"
      , legend.direction = "horizontal"
      , legend.margin = margin(0,0,0,0)
      , legend.box.margin = margin(-7,0,2,0)
      , legend.text = element_text(size = 7, angle = 25, hjust = 0.7)
      , legend.title = element_text(size = 7)
      , axis.title = element_text(size = 7)
      , plot.subtitle = element_text(size=9)
    )
```

This elevation data is cropped to the unit boundaries. Need to get elevation raster extending outside unit boundaries to avoid edge effects when aggregating HLI values to a more coarse resolution (e.g. from 1m to 30m).

Export treatment area bounding box to obtain USGS 3DEP 1m elevation data product.

```{r, eval=FALSE}
# set reproj crs
reproject_crs <- terra::crs(elev_temp)
# export
terra::ext(elev_temp) |> 
  terra::as.polygons() |> 
  sf::st_as_sf() |> 
  sf::st_set_crs(reproject_crs) |> 
  sf::st_buffer(dist = 500) |> 
  sf::st_bbox() |> 
  sf::st_as_sfc() |> 
  sf::st_write("../data/unit_bbox.shp", append = F)

```

Upload this bounding box shape to Google Earth Engine and execute [this script](https://code.earthengine.google.com/5e48227107dfd5c23f1c5b678d142a4b?noload=true) to get full USGS 3DEP 1m elevation data.

## Load Elevation Data

USGS 3DEP 1m resolution elevation data utilized for HLI calculation.

```{r}
#######################################################
#######################################################
# read elevation data
#######################################################
#######################################################


zzz <- terra::rast("../data/csf_ascc_3dep1m.tif")
zzz |> 
  as.data.frame(xy=T) |> 
  rename(elev=3) |> 
  ggplot() + geom_raster(mapping = aes(x=x,y=y,fill=elev))

ggplot() +
  geom_raster(
    data =
      zzz |>
        as.data.frame(xy=T) |>
        rename(elev=3)
    , mapping = aes(x=x,y=y,fill=elev)
  ) +
  geom_sf(
    data = 
      xxx |> 
        terra::classify(rcl = (c(-Inf, Inf, 1) %>% matrix(ncol=3, byrow=TRUE)) ) |> 
        terra::as.polygons(trunc=TRUE, dissolve=TRUE) |> 
        sf::st_as_sf() |> 
        sf::st_set_crs(reproject_crs)
    , alpha = 0, lwd = 1, color = "black"
  ) +
  scale_fill_viridis_c() +
  theme_light()
  
    

```

```{r, echo=FALSE, include=FALSE}
gc()
```


## Aspect and HLI{#hli}

Aspect of a slope (the direction or azimuth that a slope faces) strongly influences potential direct incident radiation and temperature. Untransformed, aspect is a poor variable for quantitative analysis, since 1° is adjacent to 360° – the numbers are very different even though the aspect is about the same. The following equation for Heat Load Index (HLI) rescales aspect to a scale of zero to one, with zero being the coolest slope (northeast) and one being the warmest slope (southwest). HLI is a proxy for aspect-driven solar heating.

$$
\textrm{Heat load index} = \frac{1 - \cos(\theta - 45)}{2}
$$
where $\theta$ = aspect in degrees east of north. While this and related equations are useful, they do not consider the steepness of the slope. For example, a 1° south-facing slope would receive the same heat load index as a 30° south-facing slope, even though the latter will be considerably warmer.

HLI calculations have been developed which combine slope, folded aspect (i.e., “southwestness”), and latitude to estimate terrain-driven solar heating at a given location ([McCune & Keon 2002: eq. 2](https://onlinelibrary.wiley.com/doi/10.1111/j.1654-1103.2002.tb02087.x)). To calculate HLI, we used equation 2 of [McCune & Keon 2002: eq. 2](https://onlinelibrary.wiley.com/doi/10.1111/j.1654-1103.2002.tb02087.x). Following calculation, we aggregated these data to each group selection boundary (median HLI of ~6.8m pixels). Higher values of HLI are typical on steep southwesterly slopes, while low values are common on steep northeasterly slopes. Moderate values are common in flatter terrain and on northwesterly and southeasterly aspects. 

Aspect must first be "folded" about the north-south line, rescaling 0-360° to 0-180°, such that NE = NW, E = W, etc: 

$$
\textrm{Folded aspect°} = 180° \, - \, | Aspect° – 180° |
$$

```{r}
if(file.exists("../data/topo.RDS") == TRUE & overwrite_it == FALSE){
  topo <- readRDS("../data/topo.RDS")
}else{
# extract latitude degrees from stars object
lats <- sf::st_coordinates(elev_small) %>%
  sf::st_as_sf(coords = c("x", "y"), crs = my_crs) %>%
  sf::st_transform(crs = 4326) %>%
  sf::st_coordinates() %>%
  dplyr::as_tibble() %>%
  setNames(c("longitude","latitude")) %>%
  dplyr::select(latitude)
# hli function
  # This equation returns ln(Rad, MJ · cm–2 · yr–1). It can
  # be returned to an arithmetic scale with the EXP(x)
  # function.
hli_fn <- function(slope_rad, folded_aspect_rad, latitude_rad){
  exp(
    (-1*1.236) +
      1.350*cos(latitude_rad)*cos(slope_rad) +
      (-1*1.376)*cos(folded_aspect_rad)*sin(slope_rad)*sin(latitude_rad) +
      (-1*0.331)*sin(latitude_rad)*sin(slope_rad) +
      0.375*sin(folded_aspect_rad)*sin(slope_rad)
  )
}

# calculate slope and aspect from DEM
topo <- c(
    elev_small
    , starsExtra::slope(elev_small)
    , starsExtra::aspect(elev_small)
  ) %>% 
  setNames(
    c("elev", "slope_deg", "aspect_deg")
  ) %>% 
  dplyr::mutate(
    slope_rad = pracma::deg2rad(as.numeric(slope_deg))
    , aspect_rad = pracma::deg2rad(as.numeric(aspect_deg))
    , folded_aspect_deg = 180 - abs(as.numeric(aspect_deg)-180)
    , folded_aspect_rad = pracma::deg2rad(folded_aspect_deg)
    , latitude_deg = lats$latitude
    , latitude_rad = pracma::deg2rad(as.numeric(latitude_deg))
    , hli = dplyr::case_when(
        hli_fn(slope_rad, folded_aspect_rad, latitude_rad) > 1 ~ 1
        , hli_fn(slope_rad, folded_aspect_rad, latitude_rad) < 0 ~ 0
        , TRUE ~ hli_fn(slope_rad, folded_aspect_rad, latitude_rad)
      )
  )
}
```


```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("temp_",ls())])
gc()
```


### Topography maps

```{r, eval=TRUE}
# map
vars <- c(
  "elev"
  , "slope_deg"
  , "aspect_deg"
  , "folded_aspect_deg"
  , "hli"
)
vars_title <- c(
  "Elevation (m)"
  , "Slope (\u00B0)"
  , "Aspect (\u00B0)"
  , "Folded Aspect (\u00B0)"
  , "Heat Load Index (HLI)"
)
vars_color <- c(
  "viridis"
  , "cividis"
  , "inferno"
  , "plasma"
  , "turbo"
)
vars_llim <- c(
  min(topo$elev, na.rm=T)*.95
  , 0
  , 0
  , 0
  , 0
)
vars_ulim <- c(
  max(topo$elev, na.rm=T)*1.05
  , max(as.numeric(topo$slope_deg), na.rm=T)*1.05
  , 360
  , 180
  , 1
)
# plts <- list()
plt_fn <- function(x){
  (
      ggplot() + 
      stars::geom_stars(data = (topo %>% dplyr::select(vars[x]))[]) +
      scale_fill_viridis_c(option = vars_color[x], alpha = 0.9, na.value = "transparent"
          , limits = c(vars_llim[x], vars_ulim[x])
      ) +
      geom_sf(data = units, alpha = 0, lwd = 1, color = "black") +
      labs(
          title = vars_title[x]
          , fill = vars_title[x]
        ) +
      xlab("") +
      ylab("") +
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_continuous(expand = c(0, 0)) +
      theme_bw() +
      theme(
        legend.position = "bottom"
        , legend.direction = "horizontal"
        , legend.margin = margin(0,0,0,0)
        , legend.text = element_text(size = 7)
        , legend.title = element_text(size = 7)
        , axis.text.y = element_text(size = 7)
        , axis.text.x = element_text(size = 7, angle = 35, hjust = 0.7)
        , panel.border = element_blank()
      ) 
  )
}

# combine plots
# cowplot::plot_grid(plotlist =  plts, ncol = 2)
c(1:length(vars)) %>%
  purrr::map(plt_fn)
```

## Attach topography and HLI to groups

Each group will be assigned the median value of HLI within the group boundary.

```{r}
if(file.exists("../data/all_groups_topo.gpkg") == TRUE & overwrite_it == FALSE){
  all_groups_topo <- sf::st_read("../data/all_groups_topo.gpkg")
}else{
  # functions to aggregate raster data to vector data
  topo_calc_fn <- function(dta){
    # calculate median aspect and slope by vector
    temp_slope <- aggregate(topo %>% dplyr::select(slope_rad), dta, median)  %>% 
      dplyr::as_tibble() %>%
      dplyr::select(slope_rad)
    temp_aspect <- aggregate(topo %>% dplyr::select(aspect_rad), dta, median)  %>% 
      dplyr::as_tibble() %>%
      dplyr::select(aspect_rad)
    temp_folded_aspect <- aggregate(topo %>% dplyr::select(folded_aspect_rad), dta, median) %>% 
      dplyr::as_tibble() %>%
      dplyr::select(folded_aspect_rad)
    temp_hli <- aggregate(topo %>% dplyr::select(hli), dta, median) %>% 
      dplyr::as_tibble() %>%
      dplyr::select(hli)
    # vector center
    temp_center <- sf::st_centroid(dta)$geometry %>% 
      sf::st_transform(crs = 4326) %>% 
      sf::st_coordinates() %>% 
      dplyr::as_tibble() %>% 
      setNames(c("longitude","latitude")) %>%
      dplyr::select(latitude)
    # attach to input data
    new_dta <- dta %>% 
      dplyr::mutate(
        # radians
        slope_rad = temp_slope$slope_rad
        , aspect_rad = temp_aspect$aspect_rad
        , folded_aspect_rad = temp_folded_aspect$folded_aspect_rad
        , latitude_rad = pracma::deg2rad(as.numeric(temp_center$latitude))
        # degrees
        , slope_deg = pracma::rad2deg(temp_slope$slope_rad)
        , aspect_deg = pracma::rad2deg(temp_aspect$aspect_rad)
        , folded_aspect_deg = pracma::rad2deg(temp_folded_aspect$folded_aspect_rad)
        , latitude_deg = as.numeric(temp_center$latitude)
        # hli
        , hli = temp_hli$hli
      )
    return(new_dta)
  }
  # apply to data
  all_groups_topo <- all_groups %>% topo_calc_fn()
}
```


```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("temp_",ls())])
gc()
```


## Save Data

```{r}
if(overwrite_it == FALSE){
  print("Proceeding to next section without processing or overwriting data...")
}else{
  # write data
  sf::st_write(all_groups_topo, "../data/all_groups_topo.gpkg", append = FALSE)
  sf::st_write(units, "../data/units.gpkg", append = FALSE)
  stars::write_stars(topo, "../data/topo.tif", append = FALSE)
  saveRDS(topo, "../data/topo.RDS", compress = TRUE)
}
```

